Ornaments as indicators of social changes and cultural practices in northeast Taiwan before and after European colonial period

### 2019-12-05 analysis/paper/paper.Rmd
revision based on the comments from reviewer 2

The island of Taiwan lies at the junction of mainland China, southeast Asia, and northeast Asia in the Pacific Ocean. The prehistory of Taiwan island could be roughly divided into three major periods, paleolithic (c. 27,000 BP- 5000BP), Neolithic (c. 6500- 2000BP), and Iron age (c. 2000- 400BP) with slightly regional differences in onset of each period and variations in style of artifacts and assemblages (@Liu2011; @Chen2017). It is generally accepted that Taiwan entered historical period since the early 17th century due to the colonial presence of the Spanish and Dutch who played an important role in keeping written records about Taiwan. The Spanish founded Fort San Salvador at Keelung in 1626 and Fort San Domingo in 1629 at Tamsui, and sent missionaries to local indigenous settlements in this region for religious purposes [@Blusse2000, pp. 343]. A report by a Dominican priest in 1632 mentioned that the Taparri, an indigenous tribe from northern Taiwan, exchanged carnelian beads with other indigenous groups. This form of exchange was widespread and even the Spanish soldiers learned to use carnelian as bargaining chips for gambling [@LiandWu2006, pp. 132-149]. The use of beads as prestige goods is indicated by their role in bride price payments, and compensation to resolve disputes [@LiandWu2006, pp. 132-149]. Other records mention that the female shamans (majuorbol) in the tribe would use carnelian beads as magical items in ritual practice for body healing [@Borao2009, pp. 122-151]. Records of an indigenous funeral process document the use of carnelian beads in ritual contexts, with more carnelian beads, pottery, and cloth placed into the graves of more influential people to indicate a family’s higher prestige [@LiandWu2006, pp. 153]. These historical accounts indicate that agate beads were already treated as prestige goods in Yilan before the arrival of Europeans.

@@

Chinese historical records from 1829, 1837, and 1852 during the Qing dynasty (1616-1911) contain some notes on the purposes of ornaments from Yilan [@Chen1963, pp. 228, 308; @Ke1993, pp. 11, 126; @Yao1996, pp. 77]. According to those records, indigenous people in Yilan wore ornaments in ceremonial contexts to display their wealth and status. Among those ornaments, fish-shaped necklaces made of metal threads had high value due to their delicacy and the materials invested in production. These were usually possessed by wealthy people. Other people wore agate beads or glass beads on their head or neck to participate in ceremonies. In 1895, at the beginning of Japanese colonisation, an academic field survey for plains indigenous groups reported that golden necklaces were not used in Yilan at that time, but elderly people still used beads [@Ino1996, pp. 227-232]. In sum, we find multiple sources describing how ornaments represent high status or specialised social roles in indigenous communities in Yilan. Compared to the European period, there are fewer mentions of beads in Chinese period and the descriptions are limited to clothing, but generally confirm the role of beads as status markers. However, we must recognize the inherent biases in historical records and fragmented nature of texts selected by written record keepers that might lead to misinterpretations of artifacts [@Galloway2006]. Inconsistent and incomplete descriptions from different written sources lacking contexts are the problems need to be compared and examined by studying the ornaments in archaeological contexts.

Currently the ornaments found in northeast Taiwan in early historical period around 400 hundred years ago, including glass beads, stone beads, and metal ornaments, were considered as imported materials from other regions due to a lack of evidence of beadmaking waste, metal working, or accessible local raw materials. The chemical composition analysis for glass beads shows high content of lead, coupled with the winding/folding technique, suggest a Chinese beadmaking tradition [@Cheng2008; @Gan2006; @Wang2018]. Although there is wide varity of metal ornaments such as bells, bracelets, rings, and pendants, the common compositions of metal ornaments are brass and copper with a small number made from lead and tin that indicate multiple origins mostly likely from Southeast Asia [@Chen2011]. There is no clear evidence shows the European origin of beads, however, the large amount of golden beads at Kiwulan might be related to the colonial presence of the Europeans through economic activities because similar beads were found at Luzon, northern Philippines, a part of the trading route of the Spanish between 16-19th century [@Wang2007]. Northeast Taiwan from the late Iron age to early historical period was involved in complex exchange network both in a regional and global scale.

@@

The chronology of Kiwulan can be divided into two phases represented by a upper component (700-1200 AD) and a lower component (1400-1900 AD) separated by a sterile. This chronology is based on the stratigraphy and a series of 32 radiocarbon dates previously published by @Chen2007. This paper focuses on the upper component because this has a sequence spanning pre-European contact, European presence, and Chinese presence. Previous researchers divided the upper component into six layers spanning from the 14th century to the 19th century with hundred year intervals according to the radiocarbon dates, excavation depth, consistency of contexts, and types of diagnostic porcelains such as blue and white porcelains [@Hsieh2009; @Wang2011]. Due to ambiguities in the previous chronology, I re-examined the excavation records to produce a new chronology for the sampled squares. For each sampled square I separated excavation units into three episodes of time: pre-European period, European period, and Chinese period to indicate the dominant foreign influence and cultural interaction for different time periods.

The archaeological indicators of the start of the European influence at Kiwulan are the appearance of An-ping jars and stonewares (martavans/martaban) that were introduced to Taiwan during the early 17th century. They are frequently found in European shipwrecks from this period for transporting water, wine or other fluids on the long voyages. The jar shapes found at Kiwulan are typical of those found elsewhere in VOC sites occupied during the 17th century [@Berrocal2018, pp.917; @Cort2017, pp.282; @Grave2013; @Ketel2011; @Klose2018, pp.131]. I used the presence and absence of An-ping jars and stonewares, together with the radiocarbon dates, to identify excavation units associated with the pre-European and post-European periods.

The European colonial presence in northern Taiwan ended in 1662 when the Dutch were defeated by the kingdom of Tungning founded by Koxinga from China. Later in 1683, the Qing dynasty ruled over Taiwan and a large wave of Han Chinese migrated to Yilan during the late 18th century. The archaeological signature of the Chinese period at Kiwulan is the large amount and diversity of Chinese porcelains, clay stoves. Other indicators include opium pipe-bowls and distinctive architectural bricks and tiles [@Hsieh2009]. Chinese migrations to Yilan were also recorded in official Chinese records written in the early 19th century [@Chen1963; @Ke1993].

@@

```{r tidy-data-for-table}
# clean some 
ornaments_period_clean <- 
ornaments_period %>% 
  mutate(Type = str_remove(Type, "prism/"),
         Type = ifelse(Type == "glass bead", "medium bead", Type),
         Type = ifelse(Type == "small I", "plain small", Type),
         Type = ifelse(Type == "small II", "thin small", Type),
         Type = ifelse(Type == "large thin string", "thin large", Type))
ornaments_shape <-
  ornaments_period_clean %>% 
  group_by(period, Categories, Type) %>% 
  tally() %>% 
  spread(period, n, fill = 0) %>% 
  filter(!Categories %in% c("Unknown Metal", 
                            "Perforated plate", 
                            "Knited item",
                            "Hair pin",
                            "Bone Ornament",
                            "Metal Piece",
                            "Glass ring")) %>%
  filter(!Type %in% c("other",
                      "unclassified")) %>% 
  arrange(Categories, desc(`European Presence`))
# making table
knitr::kable(ornaments_shape,
             caption = "Ornament subtype at Kiwulan. The numbers represent artefact counts ")
# some basic stat
sam_post_hole <- nrow(ornaments_period)
total_post_hole <- nrow(kwl_upper)
sample_per <- round(sam_post_hole/total_post_hole*100, 1)

@@

```{r, ornaments-image, fig.cap = "Subtypes of ornament in each major class. A: agate beads, B: bells, C: glass beads and golden beads, D: metal rings. Photographs are presented in the same order as those subtypes in the table but from left to right instead. The photographs of B, C, D classes were from original excavation report (Chen 2007).", echo=FALSE}
knitr::include_graphics(here("analysis", "figures", "ornament.png"))
```

@@

The ornaments found in the Figure \@ref(fig:plot-all-ornaments-count) shows that the most common ornament are golden beads (n = `r golden_bead_n`), followed by glass beads (n =  `r glass_bead_n`), carnelian beads (n = `r agate_bead_n`), metal rings (n = `r metal_ring_n`), and metal bells (n = `r metal_bell_n`, see also Figure \@ref(fig:ornaments-image)). The variety of subtypes and their individual frequencies are summarized in Table \@ref(tab:tidy-data-for-table). The ornaments found in the dwelling area were assumed as body adornments or a part of clothing in their daily life according to their use in the burial contexts.

@@

```{r agate-bead-fre-plot}
# plot for different shape of agate bead by period
agate_bead_shape <-
  ornaments_shape_count %>% 
  filter(Categories == 'Agate bead',
         Type != 'other')
base_size_multi_plot <- 8
agate_bead_shape_plot <- 
ggplot(agate_bead_shape) +
  geom_col(aes(reorder(Type, n), n)) +
  facet_wrap(~ period) +
  coord_flip() +
  labs(y = "",
       x = "Agate bead") +
  theme_minimal(base_size = base_size_multi_plot) +
  theme(strip.background = element_blank(), 
        strip.text = element_blank())
```

@@

```{r plot-altogether, fig.cap= "Frequency of ornament subtypes showing the changes in frequency across time periods for metal rings, agate beads, bells, and glass beads."}
library(cowplot)
plot_multi_types_by_period <- 
plot_grid(ring_shape_plot,
          agate_bead_shape_plot, 
          bell_shape_plot, 
          glass_bead_plot,
          ncol=1,
          align = "h",
          axis = "lr",
          rel_widths = 1,
          rel_heights = c(1,1,0.7,0.62))
ggsave(plot = plot_multi_types_by_period,
       here("analysis", "figures",
       "plot-multi-types-by-period.jpg"),
       width = 135, 
       dpi = 300,
       height = 95,
       units = "mm")
knitr::include_graphics(here("analysis", 
                             "figures", 
                             "plot-multi-types-by-period.jpg"))
```

@@

The distribution of frequencies for subtypes in each major class are presented in Figure \@ref(fig:plot-altogether). Spearman's correlation test shows that there is no significant relationship between diversity of subtypes and sample size (S = `r spearman_result_stat`, rho = `r spearman_result_rho`, p = $`r spearman_result_p_value`$). This indicates that the increases in diversity can be explained by the effects of culture interaction instead of effects of sample size. Carnelian beads and metal rings have greater quantity and variety of shapes compared to copper bell and glass beads during the European period. The greater varieties for carnelian beads and metal rings might indicate multiple origins due to participation in large scale trade networks stimulated by the European presence. In contrast, copper bells and glass beads have less variety that the bells are typically >2 cm long with a wide variation of human faces as a motif, while glass beads usually <0.5 cm long Indo-Pacific beads, widespread in Southeast Asian sites since 300 BC [@Francis2002].

@@

Another scenario could be that the increasing number and concentrated spatial pattern of ornaments present a form of resistance to European intrusion by using them to show social identity and emphasize the local custom had existed before European contact. Resistance could be presented in many forms, for example, @Rubertone discussed a case study in the North America that a large amount of native-made shell beads supposed to be tributes for the Europeans were found in native burial contexts. Similarly, the frequent presence of ornaments across the dwelling area not only exclusive in burial contexts suggests ornaments might be intentionally displayed in daily life to reinforce the cultural tradition. During the Chinese period, the decreasing frequency and diversity of ornaments show a decline in the production, use and discard of ornaments at Kiwulan. This reflects the smaller scale of exchange networks, with limited sources of ornaments, and limited opportunities for individuals to accumulate wealth. It is also possible that some ornaments were treated as heirlooms that passed down from one generation to another generation. However, the spacial pattern can also reflects the trend of distribution even they were heirlooms and this might suggest that the ornaments found in Chinese period might come from earlier period and continuously used till then. Further analysis about the source for each type of ornaments would be helpful to understand their origin and use context.

>>>>

The island of Taiwan lies at the junction of mainland China, southeast Asia, and northeast Asia in the Pacific Ocean. The prehistory of Taiwan island could be roughly divided into three major periods, paleolithic (c. 27,000 BP- 5000BP), Neolithic (c. 6500- 2000BP), and Iron age (c. 2000- 400BP) with slightly regional differences in onset of each period and variations in style of artifacts and assemblages (@Liu2011; @Chen2017). It is generally accepted that Taiwan entered historical period since the early 17th century due to the colonial presence of the Spanish and Dutch who played an important role in keeping written records about Taiwan. The Spanish founded Fort San Salvador at Keelung in 1626 and Fort San Domingo in 1629 at Tamsui, and sent missionaries to local indigenous settlements in this region for religious purposes [@Blusse2000, pp. 343]. A report by a Dominican priest in 1632 mentioned that the Taparri, an indigenous tribe from northern Taiwan, exchanged carnelian beads with other indigenous groups. This form of exchange was widespread and even the Spanish soldiers learned to use carnelian as bargaining chips for gambling [@LiandWu2006, pp. 132-149]. The use of beads as prestige goods is indicated by their role in bride price payments, and compensation to resolve disputes [@LiandWu2006, pp. 132-149]. Other records mention that the female shamans (majuorbol) in the tribe would use carnelian beads as magical items in ritual practice for body healing [@Borao2009, pp. 122-151]. Records of an indigenous funeral process document the use of carnelian beads in ritual contexts, with more carnelian beads, pottery, and cloth placed into the graves of more influential people to indicate a family’s higher prestige [@LiandWu2006, pp. 153]. These historical accounts indicate that carnelian beads were already treated as prestige goods in Yilan before the arrival of Europeans.

@@

Chinese historical records from 1829, 1837, and 1852 during the Qing dynasty (1616-1911) contain some notes on the purposes of ornaments from Yilan [@Chen1963, pp. 228, 308; @Ke1993, pp. 11, 126; @Yao1996, pp. 77]. According to those records, indigenous people in Yilan wore ornaments in ceremonial contexts to display their wealth and status. Among those ornaments, fish-shaped necklaces made of metal threads had high value due to their delicacy and the materials invested in production. These were usually possessed by wealthy people. Other people wore carnelian beads or glass beads on their head or neck to participate in ceremonies. In 1895, at the beginning of Japanese colonisation, an academic field survey for plains indigenous groups reported that golden necklaces were not used in Yilan at that time, but elderly people still used beads [@Ino1996, pp. 227-232]. In sum, we find multiple sources describing how ornaments represent high status or specialised social roles in indigenous communities in Yilan. Compared to the European period, there are fewer mentions of beads in Chinese period and the descriptions are limited to clothing, but generally confirm the role of beads as status markers. However, we must recognize the inherent biases in historical records and fragmented nature of texts selected by written record keepers that might lead to misinterpretations of artifacts [@Galloway2006]. Inconsistent and incomplete descriptions from different written sources lacking contexts are the problems need to be compared and examined by studying the ornaments in archaeological contexts.

Currently the ornaments found in northeast Taiwan in early historical period, including glass beads, stone beads, and metal ornaments, were considered as imported materials from other regions due to a lack of evidence of beadmaking waste, metal working, or accessible local raw materials. The chemical composition analysis for glass beads shows high content of lead, coupled with the winding/folding technique, suggest a Chinese beadmaking tradition [@Cheng2008; @Gan2006; @Wang2018]. Although there is wide varity of metal ornaments such as bells, bracelets, rings, and pendants, the common compositions of metal ornaments are brass and copper with a small number made from lead and tin that indicate multiple origins mostly likely from Southeast Asia [@Chen2011]. There is no clear evidence shows the European origin of beads, however, the large amount of golden beads at Kiwulan might be related to the colonial presence of the Europeans through economic activities because similar beads were found at Luzon, northern Philippines, a part of the trading route of the Spanish between 16-19th century [@Wang2007]. Northeast Taiwan from the late Iron age to early historical period was involved in complex exchange network both in a regional and global scale.

@@

The chronology of Kiwulan can be divided into two phases represented by a upper component (1400-1900 AD) and a lower component (700-1200 AD) separated by a sterile. This chronology is based on the stratigraphy and a series of 32 radiocarbon dates previously published by @Chen2007. This paper focuses on the upper component because this has a sequence spanning pre-European contact, European presence, and Chinese presence. Previous researchers divided the upper component into six layers spanning from the 14th century to the 19th century with hundred year intervals according to the radiocarbon dates, excavation depth, consistency of contexts, and types of diagnostic porcelains such as blue and white porcelains [@Hsieh2009; @Wang2011]. Due to ambiguities in the previous chronology, I re-examined the excavation records to produce a new chronology for the sampled squares. For each sampled square I separated excavation units into three episodes of time: pre-European period, European period, and Chinese period to indicate the dominant foreign influence and cultural interaction for different time periods.

The archaeological indicators of the start of the European influence at Kiwulan are the appearance of light grey glazed jar, known as "An-ping" jars in China and Taiwan, and large dark brown glazed stoneware jars that were introduced to Taiwan during the early 17th century. Large amount of light gray glazed jars were found on the Spanish shipwreck San Diego sinking in 1600 AD, that is believed to be the earliest evidence yet found [@Hsieh2009; @Hsieh\\]. The origin of light gray glazed jars were thought from Southeast China and commonly found at sites in Taiwan that were associated with the European activities, such as Zeelandia fort site in Tainan [@Wang2007]. Regarding large dark brown glazed stoneware jars, they might be made in Southeast Asia and are frequently found in European shipwrecks from this period for transporting water, wine or other fluids on the long voyages. The jar shapes found at Kiwulan are typical of those found elsewhere in VOC sites occupied during the 17th century [@Berrocal2018, pp.917; @Cort2017, pp.282; @Grave2013; @Ketel2011; @Klose2018, pp.131]. Those jars might be directly imported to Kiwulan by the Europeans to Kiwulan or indirectly by Chinese merchants or other indigenous groups via the regional network in north Taiwan. Either way indicates that the Europeans played an important role for introducing those jars to Taiwan. I used the presence and absence of light grey glazed jar and large stoneware jars, together with the radiocarbon dates, to identify excavation units associated with the pre-European and the start of the European influence at Kiwulan.

The European colonial presence in northern Taiwan ended in 1662 when the Dutch were defeated by the kingdom of Tungning founded by Koxinga from China. Later in 1683, the Qing dynasty ruled over Taiwan and a large wave of Han Chinese migrated to Yilan during the late 18th century. The archaeological signature of the Chinese period at Kiwulan is the large amount and diversity of Chinese porcelains in many forms such as bowls, plates, and cups. Other indicators include opium pipe-bowls and distinctive architectural bricks and tiles used by Chinese [@Hsieh2009]. Chinese migrations to Yilan were also recorded in official Chinese records written in the early 19th century recording the first immigrant in 1768 [@Chen1963; @Ke1993].

```{r C14-table-data}
C14 <- readxl::read_excel(here("analysis", "data", "raw_data", "KWL_C14_dates.xlsx"))
knitr::kable(C14,
             caption = "Radiocarbon dates on charcoals from Kiwulan")

```

@@

```{r tidy-data-for-table}
# clean some 
ornaments_period_clean <- 
ornaments_period %>% 
  mutate(Type = str_remove(Type, "prism/"),
         Type = ifelse(Type == "small bead", "small (0.5-1 cm)", Type),
         Type = ifelse(Type == "glass bead", "medium (1-2 cm)", Type),
         Type = ifelse(Type == "small I", "plain small", Type),
         Type = ifelse(Type == "small II", "thin small", Type),
         Type = ifelse(Type == "large thin string", "thin large", Type))
ornaments_shape <-
  ornaments_period_clean %>% 
  group_by(period, Categories, Type) %>% 
  tally() %>% 
  spread(period, n, fill = 0) %>% 
  filter(!Categories %in% c("Unknown Metal", 
                            "Perforated plate", 
                            "Knited item",
                            "Hair pin",
                            "Bone Ornament",
                            "Metal Piece",
                            "Glass ring")) %>%
  filter(!Type %in% c("other",
                      "unclassified")) %>% 
  arrange(Categories, desc(`European Presence`))
# making table
knitr::kable(ornaments_shape,
             caption = "Ornament subtype at Kiwulan. The numbers represent artefact counts ")
# some basic stat
sam_post_hole <- nrow(ornaments_period)
total_post_hole <- nrow(kwl_upper)
sample_per <- round(sam_post_hole/total_post_hole*100, 1)

@@

```{r, ornaments-image, fig.cap = "Subtypes of ornament in each major class. A: carnelian beads, B: bells, C: glass beads and golden beads, D: metal rings. Photographs are presented in the same order as those subtypes in the table but from left to right instead. The photographs of B, C, D classes were from original excavation report (Chen 2007).", echo=FALSE}
knitr::include_graphics(here("analysis", "figures", "ornament.png"))
```

@@

The ornaments found in the Figure \@ref(fig:plot-all-ornaments-count) shows that the most common ornament are golden beads (n = `r golden_bead_n`), followed by glass beads (n =  `r glass_bead_n`), carnelian beads (n = `r agate_bead_n`), metal rings (n = `r metal_ring_n`), and metal bells (n = `r metal_bell_n`, see also Figure \@ref(fig:ornaments-image)). The variety of subtypes and their individual frequencies are summarized in Table \@ref(tab:tidy-data-for-table). The criteria for classification of subtype is based on their dimensional attributes that are important to understand the variation in shapes. This might simplify the subtypes of glass beads when they are the same size and shape but in different color or texture. However, their variety would be considered and discussed separately. The ornaments found in the dwelling area were assumed as body adornments or a part of clothing in their daily life according to their use in the burial contexts.

@@

```{r agate-bead-fre-plot}
# plot for different shape of agate bead by period
agate_bead_shape <-
  ornaments_shape_count %>% 
  filter(Categories == 'Agate bead',
         Type != 'other')
base_size_multi_plot <- 8
agate_bead_shape_plot <- 
ggplot(agate_bead_shape) +
  geom_col(aes(reorder(Type, n), n)) +
  facet_wrap(~ period) +
  coord_flip() +
  labs(y = "",
       x = "Carnelian bead") +
  theme_minimal(base_size = base_size_multi_plot) +
  theme(strip.background = element_blank(), 
        strip.text = element_blank())
```

@@

```{r plot-altogether, fig.cap= "Frequency of ornament subtypes showing the changes in frequency across time periods for metal rings, carnelian beads, bells, and glass beads."}
library(cowplot)
plot_multi_types_by_period <- 
plot_grid(ring_shape_plot,
          agate_bead_shape_plot, 
          bell_shape_plot, 
          glass_bead_plot,
          ncol=1,
          align = "h",
          axis = "lr",
          rel_widths = 1,
          rel_heights = c(1,1,0.7,0.62))
ggsave(plot = plot_multi_types_by_period,
       here("analysis", "figures",
       "plot-multi-types-by-period.jpg"),
       width = 135, 
       dpi = 300,
       height = 95,
       units = "mm")
knitr::include_graphics(here("analysis", 
                             "figures", 
                             "plot-multi-types-by-period.jpg"))
```

@@

The distribution of frequencies for subtypes in each major class are presented in Figure \@ref(fig:plot-altogether). Spearman's correlation test shows that there is no significant relationship between diversity of subtypes and sample size (S = `r spearman_result_stat`, rho = `r spearman_result_rho`, p = $`r spearman_result_p_value`$). This indicates that the increases in diversity can be explained by the effects of culture interaction instead of effects of sample size. Carnelian beads and metal rings have greater quantity and variety of shapes compared to copper bell and glass beads during the European period. The greater varieties for carnelian beads and metal rings might indicate multiple origins due to participation in large scale trade networks stimulated by the European presence. In contrast, copper bells has less variety that the bells are typically >2 cm long with a wide variation of human faces as a motif. Although glass beads has less variety in size, small (0.5-1 cm) and medium (1-2 cm), they have wide variety of color or pattern mostly made by winding technique with high lead content in composition indicating possibly from China [@Cheng2008]. Despite more researches needed to understand the specific source of ornaments, current researches suggest that glass beads and metal ornaments have similar technique and composition to those found in China [@Chen2011; @Wang2018]. There seem to be no obvious changes in the sources of glass beads or metal ornaments at different periods in the upper component (1400-1900 AD). However, the glass beads found from lower component (700-1200 AD) are mostly Indo-Pacific beads, widespread in Southeast Asian sites since 300 BC and declined at the early 2nd millennium [@Wang2018; @Francis2002]. It indicates a different bead source more related to the exchange network in the South China Sea for the lower component.

@@

Another scenario could be that the increasing number and concentrated spatial pattern of ornaments indicate cultural resistance against the European intrusion by using them to show social identity and emphasize the local custom that had existed before European contact. Resistance could be presented in many forms. For example, @Rubertone discussed a mid-17th century Narragansett Indian cemetery where a large amount of native-made shell beads supposed to be tributes for the European were found in the graves as an act of political resistance against the colonial demands. Similarly, the frequent presence of ornaments across the dwelling area during the European suggests ornaments might be intentionally displayed in daily life to reinforce the original cultural tradition. During the Chinese period in the 19th century, the decreasing frequency and diversity of ornaments show a decline in the production, use and discard of ornaments at Kiwulan. This might reflect the smaller scale of exchange networks, with limited sources of ornaments, and limited opportunities for individuals to accumulate wealth, or population decrease. It is also possible that some ornaments were treated as heirlooms, such as carnelian beads and golden beads, that passed from one generation to the next. This might suggest that the ornaments found in Chinese period were old item continuously used till then. Further analysis about the source for ornaments from different periods will be helpful to understand their origin and reginal circulation.
